---
title: "MONet: FTICR hands on tutorial -- I"
date: "2025-02-06"
output: 
  html_document:
    code_folding: hide
---

``` {r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)

knitr::opts_knit$set(root.dir = '.')

```

This tutorial is intended to allow participants to become familiar with FT-ICR-MS data, including importing and processing, and basic exploratory visualization. 


> **USING THIS TUTORIAL**  
This .Rmd markdown file is intended to be a self-contained tutorial with text and code. 
If you downloaded the fticr_tutorial.zip file from the GitHub repo, you should have example data files in the `FTICR_data` folder. 
Load the .Rmd file in RStudio and run each code chunk to follow along. 
Or, click the "Knit" button within RStudio to generate the full report.


---

## Setup

## Load packages

You will need the `{tidyverse}` set of packages to run this tutorial. 
The `tidyverse` includes packages like `{dplyr}` and `{tidyr}` for data wrangling, and `{ggplot2}` for data visualization.


```{r setup2}

library(tidyverse)
theme_set(theme_bw(base_size = 14))

```


# I. Import and combine data

For this tutorial, we will work with the following 2 cores:

1. 60846_12
2. 60880_3

Each core was split into "TOP" and "BTM" depths, so we have a total of 4 samples for this tutorial.


Each sample was analyzed in triplicate -- all three analytical replicates are in a single zip file per sample.  
We will import all the files and combine them into a single dataframe.

```{r import}

## This script will extract all the zip files in the target folder (creating temporary files) and import and combine them into a single dataframe.
## Finally, we will delete the temporary files.



import_files = function(FILEPATH){
  
  # identify the .zip files and unzip them as csv files
  # the csv files will be saved in the parent directory (these are temporary files, will be deleted at the end)
  zip_filePaths <- list.files(path = FILEPATH, pattern = ".zip", full.names = TRUE, recursive = TRUE)
  zip_filePaths %>% lapply(unzip)
  
  # now, identify all the fticrWEOM csv files that we just extracted 
  csv_filePaths <- list.files(pattern = "fticrWEOM", full.names = TRUE)
  
  # read and combine the fticrWEOM csv files
  icr_dat <- do.call(bind_rows, lapply(csv_filePaths, function(path) {
    
    data = read.csv(path) %>% 
      mutate(source = basename(path)) # add file name
    
  }))
  
  # finally, delete the temporary files from the parent directory
  file.remove(csv_filePaths)
  
  # this is our final output
  icr_dat
}

icr_report = import_files("./fticr_data/")

print("data files imported")

```

# II. Initial cleaning

This contains a lot of info -- we do not need all these columns! In the steps below, we will select only the necessary columns.

```{r}
names(icr_report)
```


Step 1 is to split this file into two: (a) `mol` file, which has info for each molecule/peak; and (b) `dat` file, which has data about the samples


### Create molecular metadata file (`mol`)

This dataframe contains information pertaining to each peak, including the molecular formula and other molecular indices.  
We first select only the columns with atomic composition.

```{r}

mol = 
  icr_report %>% 
  dplyr::select(`Molecular.Formula`, C,H,O,N,P,S) %>% 
  rename(molecular_formula = `Molecular.Formula`) %>% 
  distinct()

names(mol)
```

Now, we want to process the `mol` file and calculate various indices

(a) indices

- AImod (aromatic index), calculated based on [Koch and Dittmar (2016)](https://doi.org/10.1002/rcm.7433)
- NOSC (nominal oxidation state of carbon), calculated based on [Riedel et al. 2012](https://doi.org/10.1021/es203901u)
- GFE (Gibbs Free Energy of carbon oxidation), calculated from NOSC, as per [LaRowe & Van Cappellen 2011](https://doi.org/10.1016/j.gca.2011.01.020)
- H/C, or the ratio of hydrogen to carbon in the molecule
- O/C, or the ratio of oxygen to carbon in the molecule

```{r}

mol = 
  mol %>% 
  mutate(across(c("N","S","P"), ~replace_na(.,0)), # convert all blank cells to 0 to help with calculations
         AImod = round((1 + C - (0.5*O) - S - (0.5 * (N+P+H)))/(C - (0.5*O) - S - N - P), 4),
         # some AImod values may be NA, or Inf, or -Inf. Change those to 0
         AImod = ifelse(is.na(AImod), 0, AImod),
         AImod = ifelse(AImod == "Inf", 0, AImod),
         AImod = ifelse(AImod == "-Inf", 0, AImod),
         NOSC = round(4 - ((4*C + H - 3*N - 2*O + 5*P - 2*S)/C),4),
         GFE = 60.3 - (28.5 * NOSC),
         HC = round(H/C, 2),
         OC = round(O/C, 2)
  )

names(mol)
```

(b) Elemental class

We can also group the molecules based on the elemental composition, i.e. "CHO", "CHONS", "CHONP", "CHONPS" classes

```{r}
mol = 
  mol %>% 
  mutate(
    # we need to drop any isotopic info (13C, 34S, etc.)
    El = str_remove_all(molecular_formula, "13C"),
    El = str_remove_all(El, "34S"),
    El = str_remove_all(El, "17O"),
    El = str_remove_all(El, "18O"),
    El = str_remove_all(El, "15N"),
    # remove any numbers from the formula to get just the elements
    El = str_remove_all(El, "[0-9]"),
    El = str_remove_all(El, " "))

names(mol)
```

(c) molecular class

Next, we assign classes (aromatic, aliphatic, etc.). These are Van Krevelen classes, typically assigned based on the H/C, O/C, and AImod indices. 
We calculate three sets of classes; users can use any of these, or assign their own classes as appropriate.

1. `Class1` from [Kim et al. 2003](https://doi.org/10.1021/ac034415p) uses H/C and O/C to classify molecules into "lipid", "unsaturated hydrocarbon", "protein", "lignin", "carbohydrate", amino sugar", "tannin", and "condensed hydrocarbon"
2. `Class2` from [Seidel et al. 2014](https://doi.org/10.1016/j.gca.2014.05.038) uses H/C, O/C, and AImod to classify molecules into "aromatic", "condensed aromatic", "highly unsaturated compounds including polyphenols/lignins", and "aliphatic"
3. `Class3` from [Seidel et al. 2017](https://doi.org/10.3389/feart.2017.00031) includes classes "aromatic", "condensed aromatic", "highly unsaturated compounds including polyphenols/lignins", "carbohydrate", "lipid", "aliphatic", and "aliphatic containing N".

```{r}
mol = 
  mol %>% 
  mutate(
    Class1 = case_when(HC >= 1.55 & HC <= 2.25 & OC >= 0 & OC <= 0.3 ~ "Lipid",
                       HC >= 0.7 & HC <= 1.5 & OC >= 0.05 & OC <= 0.15 ~ "Unsat Hydrocarbon",
                       HC >= 1.45 & HC <= 2 & OC >= 0.3 & OC <= 0.55 ~ "Protein",
                       HC >= 0.81 & HC <= 1.45 & OC >= 0.28 & OC <= 0.65 ~ "Lignin",
                       HC >= 1.48 & HC <= 2.15 & OC >= 0.68 & OC <= 1 ~ "Carbohydrate",
                       HC >= 1.34 & HC <= 1.8 & OC >= 0.54 & OC <= 0.71 ~ "Amino Sugar",
                       HC >= 0.7 & HC <= 1.3 & OC >= 0.65 & OC <= 1.05 ~ "Tannin",
                       HC >= 0.3 & HC <= 0.81 & OC >= 0.12 & OC <= 0.7 ~ "Cond Hydrocarbon",
                       TRUE ~ "Other"),
    Class2 = case_when(AImod > 0.66 ~ "condensed aromatic",
                       AImod <= 0.66 & AImod > 0.50 ~ "aromatic",
                       AImod <= 0.50 & HC < 1.5 ~ "unsaturated/lignin",
                       HC >= 1.5 ~ "aliphatic",
                       TRUE ~ "other"),
    Class3 = case_when(AImod > 0.66 ~ "condensed aromatic",
                       AImod <= 0.66 & AImod > 0.50 ~ "aromatic",
                       AImod <= 0.50 & HC < 1.5 ~ "unsaturated/lignin",
                       HC >= 2.0 & OC >= 0.9 ~ "carbohydrate",
                       HC >= 2.0 & OC < 0.9 ~ "lipid",
                       HC < 2.0 & HC >= 1.5 & N == 0 ~ "aliphatic",
                       HC < 2.0 & HC >= 1.5 & N > 0 ~ "aliphatic+N",
                       TRUE ~ "other")
  )
         

print("`mol` created")

```


this file now contains most of the info needed to interpret the data across different samples.

```{r}
names(mol)
```

---

### Create `dat`

This dataframe contains info about the samples being analyzed, i.e., which peaks were identified in which sample.  
We will clean this up and eventually merge with the `mol` dataframe.

```{r}

dat = 
  icr_report %>% 
  dplyr::select(source, Molecular.Formula, Calculated.m.z, contains("Peak.Area")) %>% 
  janitor::clean_names() %>% 
  separate(source, sep = "_", into = c("icr", "Proposal_ID", "Sampling_Set", "Core_Section", "Rep")) %>% 
  mutate(Rep = parse_number(Rep),
         sample_name = paste0(Proposal_ID, "_", Sampling_Set, "_", Core_Section)) %>% 
  dplyr::select(-icr)

```


#### 3 acquisitions

Each sample/rep was analyzed 3 times on the machine (i.e., 3 acquisitions, or instrument replicates).  
For robust analysis, some users may choose to only include peaks identified across multiple acquisitions (e.g., peaks seen in 2 of 3 acqusitions, or peaks seen in all acquisitions). The choice is dependent on the user and the questions being asked.  
For this example, we only include peaks identified in 2 of the 3 acquisitions. 


```{r}
## acquisition reps

# an easy way to do this is to convert all the peak areas to binary 0/1 and then add them up. 
# if the sum is 2 or more, that means the peak was seen in 2 or more acquisitions

dat_acq = 
  dat %>% 
  mutate(peak_area_1 = case_when(peak_area_1 > 0 ~ 1),
         peak_area_2 = case_when(peak_area_2 > 0 ~ 1),
         peak_area_3 = case_when(peak_area_3 > 0 ~ 1),
         acquisition_count = peak_area_1 + peak_area_2 + peak_area_3,
         acquisition_KEEP = acquisition_count >= 2) %>% 
  filter(acquisition_KEEP) %>% 
  dplyr::select(-c(contains("peak_area"), contains("acquisition")))

```


#### 3 replicates

Each sample was extracted 3 times. We apply the same 2/3 replication filter as above. This is also user-dependent.

```{r}
## Now, we only select molecules that were identified in 2/3 of the total reps

# identify the number of replicates for each sample
# for this MONet workflow, each sample has 3 replicates, so it is straightforward.
# But this code can be used even if there was uneven replication across different samples.

max_replicates = 
  dat_acq %>% 
  dplyr::select(sample_name, Proposal_ID, Sampling_Set, Core_Section, Rep) %>% 
  distinct() %>% 
  group_by(sample_name, Proposal_ID, Sampling_Set, Core_Section) %>% 
  dplyr::summarise(total_reps = n()) 

# now, calculate the occurrence of each peak 
dat_reps = 
  dat_acq %>% 
  left_join(max_replicates) %>% 
  group_by(Proposal_ID, Sampling_Set, Core_Section, molecular_formula) %>% 
  dplyr::mutate(peak_reps = n()) %>%
  ungroup() %>% 
  mutate(KEEP = peak_reps >= (2/3) * total_reps) %>% 
  filter(KEEP) %>% 
  dplyr::select(-KEEP)

dat_final = 
  dat_reps %>% 
  dplyr::select(-c(Rep, total_reps, peak_reps)) %>% 
  distinct()

print("`dat_final` created")

```

This is the list of all the peaks "present" (identified) in our samples.

Now, combine this file with the `mol` file we generated above.


```{r}
icr_processed = 
  dat_final %>% 
  left_join(mol)

print("`icr_processed` created")

names(icr_processed)
```

---- 

# III. Summary of the data

Number of peaks identified in each sample:

```{r}
icr_processed %>% 
  group_by(sample_name) %>% 
  dplyr::summarise(count = n()) %>% 
  knitr::kable()
```


Grouped by element composition

```{r}
icr_processed %>% 
  group_by(sample_name, El) %>% 
  dplyr::summarise(count = n()) %>% 
  pivot_wider(names_from = "El", values_from = "count") %>% 
  knitr::kable()
```

Grouped by compound class

```{r}
icr_processed %>% 
  group_by(sample_name, Class2) %>% 
  dplyr::summarise(count = n()) %>% 
  pivot_wider(names_from = "Class2", values_from = "count") %>% 
  knitr::kable()
```

<br>

<br>

We can also plot histograms to determine spread of the data

```{r, fig.width=8, fig.height=3}

icr_processed %>% 
  ggplot(aes(x = calculated_m_z, color = Core_Section))+
  geom_histogram(position = "identity", fill = NA, linewidth = 1)+
  facet_wrap(~Proposal_ID)

icr_processed %>% 
  ggplot(aes(x = NOSC, color = Core_Section))+
  geom_histogram(position = "identity", fill = NA, linewidth = 1)+
  facet_wrap(~Proposal_ID)

```


---

# IV. Van Krevelen Plots and Molecular Classes

Van Krevelen plots are a way to visualize the molecular composition of the samples.  
All peaks are plotted as a function of H/C ~ O/C, and the different regions are assigned molecular classes. 

Here are some examples, based on the classification systems used above.

### VK Domains

```{r, fig.height=12, fig.width=17}
vk_domains = 
  icr_processed %>% 
  dplyr::select(HC, OC, Class1, Class2, Class3)

vk1 = 
  vk_domains %>% 
  ggplot(aes(x = OC, y = HC, color = Class1))+
  geom_point()

vk2 = 
  vk_domains %>% 
  ggplot(aes(x = OC, y = HC, color = Class2))+
  geom_point()

vk3 = 
  vk_domains %>% 
  ggplot(aes(x = OC, y = HC, color = Class3))+
  geom_point()

cowplot::plot_grid(vk1, vk2, vk3)
```


## VK Patterns in samples

We now apply these VK plots to see the peaks identified in our samples

```{r}

icr_processed %>% 
  ggplot(aes(x = OC, y = HC))+
  geom_point(size = 0.5)+
  facet_wrap(~ sample_name)

```

Samples identified by color

```{r}

icr_processed %>% 
  ggplot(aes(x = OC, y = HC, color = sample_name))+
  geom_point(size = 0.5)

```

Because of the sheer number of peaks present, it is difficult to identify differences among samples.  
One solution is to identify the unique peaks present in each sample and plot those. 

## Unique peaks

Q: which peaks are unique to TOP vs. BTM sections in each core?

```{r, fig.width=12}
# to identify unique peaks,
# compute how many times a peak is seen across the groups
# a count of 1 means the peak is unique to that sample

unique_top = 
  icr_processed %>% 
  group_by(molecular_formula, Proposal_ID) %>% 
  dplyr::mutate(count = n()) %>% 
  ungroup()

unique_top %>% 
  filter(count == 1) %>% 
  ggplot(aes(x = OC, y = HC, color = Core_Section))+
  geom_point(size = 2)+
  facet_wrap(~ Proposal_ID + Sampling_Set) +
  labs(title = "Unique Peaks",
       color = "Unique to:")

```


<br>

Q: How do the two sites compare to each other? What are the unique peaks in 60880 vs. 60846 TOP soils?

```{r}

unique_site = 
  icr_processed %>% 
  filter(Core_Section == "TOP") %>% 
  group_by(molecular_formula) %>% 
  dplyr::mutate(count = n())

unique_site %>% 
  filter(count == 1) %>% 
  ggplot(aes(x = OC, y = HC, color = Proposal_ID))+
  geom_point(size = 2)+
  stat_ellipse(level = 0.90, linewidth = 1, aes(group = Proposal_ID), color = "black")+
  labs(title = "Unique Peaks",
       color = "Unique to:")

```

Here, we have added the density ellipse, which shows us the region where 90 % of the peaks are seen for each sample. From the graph, it is clear that the two samples contain different types of molecules.


--- 


# V. Relative abundance

Another metric we use is the "relative abundance" or "relative contribution". This allows us to see the general distribution of the different molecular classes across the samples.


```{r}

rel_abundance = 
  icr_processed %>% 
  group_by(Proposal_ID, Sampling_Set, Core_Section, sample_name, Class2) %>% 
  dplyr::summarise(count = n()) %>% 
  group_by(Proposal_ID, Sampling_Set, Core_Section, sample_name) %>% 
  dplyr::mutate(total = sum(count),
                relabund = 100 * count/total,
                relabund = round(relabund, 2)
                ) %>% 
  ungroup()

rel_abundance %>% 
  dplyr::select(sample_name, Class2, relabund) %>% 
  pivot_wider(names_from = "Class2", values_from = "relabund") %>% 
  knitr::kable()

```

<br>

We can plot this using stacked bar graphs

```{r}
rel_abundance %>% 
  ggplot(aes(x = Core_Section, y = relabund, fill = Class2)) +
  geom_bar(stat = "identity")+
  facet_wrap(~Proposal_ID)
```


--- 
---

